---
title: "A fistful of ducats"
author: "Einar"
date: 2020-11-30
description: "Some notes on utility and insurance"
#categories: ["R"]
tags: ["visualization", "utility", "insurance", "ducats"]
math: true
editor_options: 
  chunk_output_type: console
---
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, dev.args=list(bg="transparent"))
knitr::opts_knit$set(global.par=TRUE)
```

```{r, echo=F}
# Do this once for all plots
par(cex.lab=1.25, cex.axis=1.25, cex.main=1.25, cex.sub=1.25, cex= 1.3, 
    lwd=2, pch=20, bty="n")
```
Some notes on utility and insurance. The below is apropos of ET Jaynes'
_Probability Theory_, Chapter 13. It's a useful and interesting book, but Jaynes
was apparently no great believer in graphical illustrations.

## Utility functions
Daniel Bernoulli, one of the many Bernoullis, conceived a utility function 
$u(x) \propto log(x)$ in contrast to the linear $u(x) \propto x$, usual at the time. 
Naturally $x$ denotes money: Gambling was the principal object of probability
arguments in the mid-1700s. Utility is hazy, but let us say that it measures the
usefulness of [a ducat](https://en.wikipedia.org/wiki/Ducat) in the unit
_utils_. Daniel called it the "moral value" of money. Linear utility
tells us that earning another ducat is always useful and never gets less useful
no matter how many you have. The Scrooge McDuck idea of money. Daniel's
logarithmic utility encodes the common sense that someone with 1000 ducats has
less use for another ducat than someone with only five ducats.

If you gamble to maximize expected profit as per linear utility you may end up 
with some
[nonsense betting strategies](https://en.wikipedia.org/wiki/St._Petersburg_paradox). 
Suppose your fortune is $M$ ducats. One month's salary, let's say. You are to
bet $m \leq M$ ducats on the flip of a coin, with a payout of $2m$ if you win.
How much should you bet if you are pretty sure there is a $.51$ probability of
heads? Your expected fortune after playing this game is 
$$\mathbb E(\mathrm{fortune}) = .51(M + m) + .49(M - m) = M + .02m.$$
To maximize this simply take $m$ as large as possible. Ask your mum for a loan.

Despite this compelling evidence I'm not sure I would bother betting any
ducats on this game. Perhaps my utility is logarithmic? If so it grows quickly
in the beginning and more slowly later on. This implies that the losing of a
ducat causes me more pain (measured by lost utils) than the winning of a ducat
gives me pleasure. Especially at the lower end of
the scale where there is larger curvature. An example (plot below): If I own a
single ducat, the "moral value" of my fortune is $\log(1) = 0$ utils.  If
I lose half a ducat it will cost me about $.7$ utils. But if I win half a ducat
it will gain me only $.4$ utils!

```{r, echo=F}
curve(log(x), from=0, to = 2, 
      xlim=c(.5,1.5),
      ylim=c(log(.5), log(1.5)),
      main="When all you have is a ducat",
      xlab="ducats",
      ylab="utils")

lines(c(1, 1), c(log(1), log(1.5)), col="deepskyblue4", lwd=1.5)
lines(c(1, 1), c(log(.5), log(1)), col="goldenrod", lwd=1.5)
lines(c(1, 1.5), c(log(1.5), log(1.5)), lwd=1.5, lty=3, col="deepskyblue4")
lines(c(1, .5), c(log(.5), log(.5)), lwd=1.5, lty=3, col="goldenrod")
points(1, 0, pch=20)
points(.5, log(.5), pch=20, col="goldenrod")
points(1.5, log(1.5), pch=20, col="deepskyblue4")
text(1.25, -.4, "utils lost by \n losing a half-ducat", col="goldenrod")
text(.75, .25, "utils gained by \n gaining a half-ducat", col="deepskyblue4")
```

These numbers have no direct interpretation as far as I know, but we can see some interesting relations: Logarithmic utils are proportional to the number of digits of your fortune. And since $\log(ax) = \log(a) + \log(x)$ we can say that my loss of a half-ducat pains me as much as she who has ten ducats and loses five. She counts single ducats in the same way that I count tenths-of-ducats. We are the same but for order of magnitude.

With logarithmic utility I might still be willing to bet some amount, $m$, on the $.51$ chance of heads. Per the proportionality argument above my fortune of a single ducat, $M=1$, can be interpreted as "your entire fortune whatever it is," and you can simply multiply it up. Maybe still a month's salary. Now $m$ is the _proportion_ of your fortune you'd be willing to bet. We want to maximize expected utility 
$$\mathbb E(\mathrm{utility}) = .51\log(1 + m) + .49\log(1 - m).$$
This is maximum at roughly $m=.02$, shown below, but the difference in utility between this and simply not betting is negligible. Going by the monthly salary of the average Norwegian this bet could buy you a mid-range bottle of whisky (Norwegian prices, probably several bottles in another country). I'd rather just take the bottle. We can also see by the increasing steepness of this curve that I'd be an idiot to bet my entire ducat. Unless the thrill of betting is in itself _very_ valuable to me.

```{r, echo=F}
curve(.51*log(1 + x) + .49*log(1-x), from=0, to = 1, 
      ylim=c(-.5, 0),
      ylab="E(utils)", xlab="Size of bet",
      main ="Expected utility with .51 win probability")
mx <- optimize(function(x) .51*log(1 + x) + .49*log(1-x), 
               c(0,1), maximum = T)$maximum

abline(h=0, lty=3, col="grey", lwd=1)
points(mx, .51*log(1 + mx) + .49*log(1-mx), pch=20)
```

What if we sweeten the pot by increasing the probability of heads to $.7$? Now my expected utility (see below) is clearly increased if I bet something. It is maximized somewhere around $.4$. The function has an interesting feature in that I should feel the same about betting nothing at all and betting something like $.7$. Does this correspond to common sense? It's not entirely far-fetched to me. At such good odds I would feel uncomfortable betting nothing; but I would also feel uncomfortable betting the larger part of my fortune. Less than half feels OK.
```{r, echo=F}
curve(.7*log(1 + x) + .3*log(1-x), from=0, to = 1, ylab="E(utils)", xlab="Size of bet",
      main ="Expected utility with .7 win probability", ylim=c(-.4, .4))
mx <- optimize(function(x) .7*log(1 + x) + .3*log(1-x), c(0,1), maximum = T)$maximum

points(mx, .7*log(1 + mx) + .3*log(1-mx), pch=20)
abline(h=0, lty=3, col="grey", lwd=1)
```

## Why do I buy insurance?
Insurance is a racket if you have linear utility. The insurance sellers must
necessarily demand premiums that are larger than the expected payouts. How could
they not? They'd be giving ducats away for free. At the same time why should I
pay more than my expected payout? I'd be giving ducats away for free.

One answer[^1] lies in the difference between my utility and that of the
insurance seller. Under logarithmic utility the insurance seller
and I operate on very different scales. The loss of a half-ducat is pretty bad
to me, but the insurance seller who has, say, 1000 ducats, will not notice such
a loss. In fact, on the scale of half-ducats, the insurance seller has more or
less linear utility:
```{r, echo=F}
curve(log(x), from=998, to = 1002, 
      xlim=1000+c(-1,1),
      ylim=c(log(1000-0.5), log(1000 + .5)),
      main="When you have 1000 ducats",
      xlab="ducats",
      ylab="utils")

lines(c(1000, 1000), c(log(1000), log(1000.5)), col="deepskyblue4", lwd=1.5)
lines(c(1000, 1000), c(log(999.5), log(1000)), col="goldenrod", lwd=1.5)
lines(c(1000, 1000.5), c(log(1000.5), log(1000.5)), lwd=1.5, lty=3, col="deepskyblue4")
lines(c(1000, 999.5), c(log(999.5), log(999.5)), lwd=1.5, lty=3, col="goldenrod")
points(1000, log(1000), pch=20)
points(999.5, log(999.5), pch=20, col="goldenrod")
points(1000.5, log(1000.5), pch=20, col="deepskyblue4")
text(1000.45, log(999.75), "utils lost by \n losing a half-ducat", col="goldenrod")
text(999.5, log(1000.25), "utils gained by \n gaining a half-ducat", col="deepskyblue4")
```

This is basic calculus: if you zoom in far enough on a curve it's going to look
straight. Imagine I will have an unlikely but expensive mishap at the
flip of a coin. Heads, with $.01$ probability, I will need to pay $.99$ of my
single ducat; tails nothing happens. With no insurance my expected utility is
$$.01\log(1 - .99) + .99\log(1) \approx -.05.$$ 
The expected payout is $.01*.99 + .99*0 \approx .01$, so if I pay exactly this
premium (and the insurer earns nothing on average), my expected utility is
$$.01\log(1 - .01) + .99\log(1-.01) = \log(1-.01) \approx -.01,$$
which is better for me than not having insurance.

[^1]: Another answer, left as an exercise to the reader, has to do with variance. This argument may also apply to flipping coins with linear utility.

The break-even point, where insurance isn't worth it for me in terms of utility,
is the premium $P$ where $\log(1-P) = -.05$. For small $P$ we have 
$\log(1-P) \approx -P$, so the break-even point is around $.05$. In principle the
insurance seller and I can both be happy if the premium is somewhere in 
$[ .01, .05 ]$.

## All models are wrong
You can get nonsense strategies by adhering to logarithmic utility, too. The
"moral value" of money might not be a simple mathematical function and it might
be different for different people. Personally, I think there is some threshold
above which I would lose all interest in gaining ducats. Jordan Ellenberg, in
his book _How Not to be Wrong,_ describes a real-life example of such a bounded
utility function: Writer Fran Lebowitz would, in her youth, start the month
driving a cab until she had made enough money to cover rent and food. After this
she would stop driving the cab and spend the rest of the month writing.
